---
title: "MongoDB & R"
author: "Vik & Gaurav"
date: "16 March 2016"
output: 
  ioslides_presentation: 
    smaller: yes
    transition: faster
    widescreen: yes
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyjson)
library(rjson)
library(rmongodb)
library(dplyr)
```

## Connecting to MongoDB

As long as you have your MongoDB instance running, it is relatively 
painless to connect from R:

```{r connect_1}
mongo <- mongo.create(host="127.0.0.1",db="test")
mongo.is.connected(mongo)
```

    
# Analysis 0: Grades

## Restaurant Names and Grades

Suppose that we wish to obtain a data frame consisting of all restaurants 
and their grades.

Since we have a small dataset, let's read it all in and store it in R:
```{r read_all}
all_data <- mongo.find.all(mongo, "test.restaurants")
length(all_data)
```

## Output Format (all_data)

The output format is a list, with one component for each document.

```{r read_all_2}
class(all_data)
str(all_data[[1]], max.level=1)
```

## Attempt 1: Unlisting Directly

"do.call" Method
```{r do_call_1, warning=FALSE, echo = TRUE, cache=TRUE}
r1 <- tryCatch ( 
  { do.call("rbind", lapply(all_data, data.frame)) },
  error = function(cond){
  print(cond)
  return(NULL)
  }
)
```
What causes this method to fail?
- Uneven number of columns per restaurant. Due to "grades" data 

## Distribution of Number of Grades

```{r grades_pic}
num_g <- sapply(all_data, function(x) length(x$grades))
barplot(table(num_g), col="lightsalmon3", border=NA, 
        xlab="Number of gradings", ylab="Number of restaurants")
abline(h=seq(0, 5000, by=1000), col="white", lwd=0.4, lty=2)
```

## Fix 1: Using a `for` Loop

```{r fix_1, warning=FALSE, echo = TRUE}
nm <- vector(mode="character", length = length(all_data))
cusne <- vector(mode="character", length = length(all_data))
grds <- vector(mode="character", length = length(all_data))
for(a in 1:length(all_data)) {
  nm[a] <- all_data[[a]]$name
  cusne[a] <- all_data[[a]]$cuisine
  grd <- sapply(all_data[[a]]$grades, function(y) y$grade)
  grds[a] <- paste0(grd, collapse =  "")
}
r <- data.frame(name = nm, cuisine = cusne, grades=grds)
glimpse(r)
```

## Fix 2: Using `lapply`
```{r fix_2, warning=FALSE, echo = TRUE}
get_rest_dtls <- function(x){ 
  nm <- x$name
  cusne <- x$cuisine
  grds <- sapply(x$grades, function(y) y$grade)
  grades <- paste0(grds, collapse =  "")  
  r <- data.frame(name = nm, cuisine = cusne, grades = grades,
                  stringsAsFactors = FALSE)
}
r1 <- lapply(all_data,get_rest_dtls)
r2 <- rbind_all(r1)
glimpse(r2)
```

## Fix 3: Using `tidyjson`

```{r fix_3}
# extract only grades and name
f2 <- list("name"=1L, "cuisine"=1L, "_id"=0L, "grades"=1L)
out <- mongo.find.all(mongo, "test.restaurants", fields=f2)

# convert to tidyjson class
out.ch <- sapply(out, rjson::toJSON)
out2 <- as.tbl_json(out.ch)

# extract grades for each restaurant
out2 %>% spread_values(name=jstring("name")) %>% 
  enter_object("grades") %>% gather_array() %>% 
  enter_object("grade") %>% append_values_string("grade") -> out2a
head(out2a)
```

# Analysis 1: Above B

## Good Restaurants

Suppose that we wish to obtain a data frame consisting of all restaurants 
that have never had a grade below B.

The unique grades are 
```{r unique_grades}
unique(out2a$grade)
```

## Using Ordered Levels

```{r analysis_2}
out2b <- out2a
out2b$grade <- factor(out2b$grade, ordered=TRUE)
levels(out2b$grade)

# write a helper function
id_poor <- function(tmp) {
  data.frame(poor=any(tmp$grade > "B"))
}
out2b %>% group_by(name) %>% do(id_poor(.)) -> poor_good
```

## Double Check Results
```{r analysis_2a}
head(poor_good, n=3)
dplyr::filter(out2b, name == "#1 Sabor Latino Restaurant")
```
    
# References

## References

```{r mongo_out, echo=FALSE}
mongo.destroy(mongo)
```


1. Getting Started with NoSQL, *Gaurav Vaish*
2. `rmongodb` vignette
3. `tidyjson` vignette
  

    