---
title: "Non-standard Evaluation"
author: "Vik"
date: "23 November 2015"
output: 
  ioslides_presentation: 
    smaller: yes
---

## What does NSE refer to?

```{r nse_0, echo=FALSE, message=FALSE}
library(dplyr)
```

Consider the following dataset.
```{r nse_1}
head(mtcars)
ls()
```
Notice that my global environment (or my workspace) is empty.

## Examples of Non-Standard Evaluation
Consider the following codes
```{r nse_2}
head( dplyr::filter(mtcars, hp <= 100), n=3)
head( subset(mtcars, cyl == 8), n=3)
```

* Recall that our global workspace was empty. Where did R know to find the variables 
`hp` and `cyl`?
* Now let's write a similar function of our own.

## Why Can't I do That?

```{r nse_3}
f <- function(col) {
  dplyr::filter(mtcars, col <= 100)
}
# Neither of this will work:
# f(hyp)
# f("hyp")
```

* The problem is that the expression `col1 <= 100` is not being evaluated correctly.
* We need to revert to a standard evaluation approach now and do the following:
    * construct the expression we want.
    * assist R in evaluating it.

## Using SE Versions of dplyr verbs I
```{r nse_4}
f <- function(col, f_lim=100) {
  filter_crit <- lazyeval::interp(~ var1 <= f_lim, var1=as.name(col))
  filter_(mtcars, filter_crit)
}
head( f("hp", 100), n=2)
head( f("mpg", 20.0), n=2)
```

## Using SE Versions of dplyr verbs II
```{r nse_5}
f <- function(col, f_mult=0.45) {
  mutate_crit <- lazyeval::interp(~ var1 * f_mult, var1=as.name(col))
  mutate_(mtcars, mutate_crit)
}
head( f("mpg", 1.61), n=2)
head( f("wt", 0.45), n=2)
```

## Using SE Versions of dplyr verbs III
```{r nse_6}
f <- function(col, f_mult=0.45, new_col_name) {
  mutate_crit <- lazyeval::interp(~ var1 * f_mult, var1=as.name(col))
  mutate_(mtcars, .dots = setNames(list(mutate_crit), new_col_name))
}
head( f("mpg", 1.61, "kmpg"), n=2)
```

## Who Will Find This Most Useful?

* People who need to use `dplyr` verbs within a function.
* Package developers, whose code imports `dplyr`.

## Helpful links:

* [Hadley Wickham's page](http://adv-r.had.co.nz/Functional-programming.html) on NSE.
* [How R searches and finds stuff](http://blog.obeautifulcode.com/R/How-R-Searches-And-Finds-Stuff/)
* The dplyr vignette on NSE: 
```{r eval=FALSE}
library(dplyr)
vignette("nse")
```
* [R bloggers page](http://www.r-bloggers.com/using-mutate-from-dplyr-inside-a-function-getting-around-non-standard-evaluation/) on NSE
